Analysis of observational data, to calculate the Delayed-Tau model parameters

### 2025-03-26 thesis.qmd
minor pdf fixes

abstract: |
  The Star Formation History (SFH) of a galaxy can offer many insights not only for the evolution and the future of the galaxy, but also for the evolution of the Universe. This is why there are various theoretical models trying to describe the SFH of galaxies. One of those models is the Delayed-Tau model, which approximates the Star Formation Rates (SFR) of galaxies as a function with a rising SFR at the beginning, until it reaches a peak at a time $\tau$, unique for each galaxy, and then it drops at an exponential rate.

  @haslbauerCosmologicalStarFormation2023 argue that the use of Delayed-$\tau$ model for the galaxies of the Local Cosmological Volume (LCV) is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observatory data for SFR and Stellar Masses from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LV), allows the investigation of the SFR throughout the life of each galaxy and so we can find the expected time of peak of the SFR.

@@

# Galaxy Morphology and Star-Forming Regions

@@

[^1]: The chain is homogeneous if $T \forall x^{(i)}$ remains invariant (the probability of moving from $x^{(i−1)}$ to $x^{(i)}$ does not depend on the iteration number $i$.) and $\sum_{x^{(i)}} T\bigl(x^{(j)} \mid x^{(i-1)}\bigr) = 1$.

@@

    -   Assigned positive T-values , extending to +11 for extreme irregulars.

@@

@@

# Star Formation History (SFH)

@@

$$
\int_0^{t_{sf}} \text{SFR}(t)dt = \zeta M_*(t_{sf}),\ t_{sf}=\text{Time of Star Formation,} 
$$ {#eq-sfr_int}

@@

-   **H**\boldmath$\alpha$ **Emission**: Young, hot, massive stars (O-type stars, \~10 Myr, \~20 $M_\odot$) produce a number of ionizing photons, which they ionize the surrounding hydrogen rich gas. The hydrogen undergoes recombination cascades which produce Balmer emission lines of $H\alpha$ (0.6563 $\mu m$) and $H\beta$ (0.4861 $\mu m$). Dust can significantly affect observations.

-   **Far-Ultraviolet (FUV) Flux**: Mainly emitted by young, hot stars ( B-type stars, \~100 Myr). Dust presence can also significantly affect observations.

-   **Infrared (IR) Flux**: The stars in a galaxy can heat up the dust in different ways, which then emits radiation in different parts of the IR spectrum. For example, young and massive, short-lived stars, emit UV radiation which then the heated dust emits in a wavelength of $\approx 60 \mu m$, whereas dust heated by UV-faint old or low-mass stars will emit at $\approx 100-150\mu m$. As a result, the total IR emission is age-agnostic and provides a more accurate approximation of the SFR because it accounts for contributions from both young and old stellar populations.

-   **Radio Continuum Emission**: Strongly correlated with IR. Its origin is complex, involving synchrotron radiation from relativistic electrons and thermal Bremsstrahlung from hot gas.

-   **X-Ray Emission**: In star-forming galaxies, X-rays arise from high-mass binary systems (neutron star or black hole with massive stellar companion) and hot gas from supernovae, correlating with SFR up to redshift $z \sim 4$. X-rays are dust-insensitive, enabling accurate high-redshift observations.

@@

where $\mathcal{K}_i$ is a constant specific to each $L_i$ ($i =$ H$\alpha$, IR, radio, FUV, X). In our analysis, we lack radio and X-ray data.

@@

![Plot showing the linear relation $\log\text{SFR}_{FUV} = \log\text{SFR}_{H\alpha}$, as well as their distributions, based on the data from UNGC](figure/log_SFR_FUV_Ha.png){#fig-SFR-Ha-FUV fig-align="center" width="500"}

@@

where $\alpha(t)$ and $\beta(t)$ depend on time and redshift $z$ (@speagleHighlyConsistentFramework2014):

### Star Formation History Models

![Star Formation Rate over the time of star formation, for different parametric models](images/sfr-models.png){fig-align="center" width="300"}

@@

@@

![The history of cosmic star formation as shown in @madauCosmicStarFormation2014](images/lily-madau.png){fig-align="center" width="450"}

@@

@@

where $\text{SFR}_0$ is given in the catalogs. According to these model the timescale, at which the SFR peaks, is around $3.5\lessapprox\tau\lessapprox 4.5$ [@speagleHighlyConsistentFramework2014]. If we intergrate the SFR(t), we can calculate the present day averaged SFR:

@@

$$
\overline{\text{SFR}_\text{del}} = \frac{\zeta M_*}{t_sf}
$$ {#eq-averaged-SFR-M}

@@

In our current analysis, we will use two different catalogs of galaxies, containing galaxies within the Local Cosmological Volume (LV), which is defined as a sphere with a radius of 11 Mpc centred around the Milky Way, to remain consistent with with the analysis of @kroupaConstraintsStarFormation2020 and @haslbauerCosmologicalStarFormation2023. The reason we focus on the LV and thus only at small distances is because we want to understand the SFR of the galaxies as they are today and since the redshift depends on distance, we can assume that the redshift of our galaxy sample is $z = 0 \Rightarrow t_\text{emitted}=T_0 = 13.8$ Gyr.

@@

-   **Luminosity and Distance Dependence:** The HECATE catalog is more complete for brighter galaxies (with luminosities above $10^{9.5} L_{B,\odot}$) and distances within 33 Mpc, which includes the LV. For galaxies with luminosities around $10^{10} L_{B,\odot}$, the catalog remains complete up to distances of 100 Mpc.
-   **Luminosity Density:** The catalog's completeness is also estimated based on the luminosity density in the B-band. Within 30 Mpc, the catalog is over 100% complete due to the high density of galaxies around the Milky Way. However, at distances around 170 Mpc, the completeness drops to approximately 50%, indicating that many galaxies are missing.
-   **Star Formation Rate and Stellar Mass:** The completeness of SFR is around 50% for distances between 30 and 150 Mpc, primarily due to limitations in the WISE survey coverage. Similarly, the completeness of stellar mass estimates is comparable to that of the B-band luminosity, with overdensities at short distances and a sharp cutoff at larger distances.

The *UNGC* catalog focuses on galaxies within the LV, specifically those with radial velocities less than 600 km/s and distances within 11 Mpc. It is around $40\%-60\%$ complete for galaxies with B-band magnitude $M_B^c < -11^m$. In the Local Group ($D \leq 10 \, \text{Mpc}$) only half the galaxies have $M_B^c > -11^m$ and it is expected that for distances $\sim 9 \, \text{Mpc}$ more than half of the ultra-faint galaxies are missing. Specifically, it is estimated that around $10^3–10^4$ galaxies of the LV are not included in UNGC.

@@

Since convergence can not occur for $x\ll 1$, this method should not be used for galaxies in the begging phases of their star formation [^10]. If the assumption, $t_{sf}=13.6$ Gyr for all the galaxies is correct, then an inconstancy in the plot $A_{del}-\text{sSFR}$ should only be observed, for higher values of the sSFR, since, according to the delayed- $\tau$ model, higher sSFR values indicate a rising SFR, while lower SFR values indicate that the bulk of $M_*$ is already produced and the galaxy is in the declining SFR phase. However, as seen from @fig-sSFR-A, this inconsistency occurs in galaxies for both high and low sSFR's, meaning that the the model has convergence issues both for slow rising SFR and quenching galaxies, indicating an intrinsic problem with this approach.

@@

Using these priors, the metrics of the Stan program give good results ($\hat{R}<1.001$ and $\text{ESS} >1000$), while the calculated values correspond to the data of the catalog. Specifically, we use the data to recalculate the SFR from the predicted parameters, using the delayed-$\tau$ model and compare the normalization constant to the stellar masses of the galaxies (@fig-A-M-mcmc).

::: {#fig-distr-comparison layout-ncol="2" fig-pos="b"}
![Distribution of $\tau$ values.](method_comparison/tau_hist.png){#fig-tau-hist-both}

@@

@@

@@

::: {#fig-A-M-mcmc layout="[[50,50],[1]]" fig-pos="hb"}
![Uniform $\tau$ prior](method_comparison/A_M_MCMC_uni.png){#fig-A-M-mcmc-uni}

@@

![Main Sequence of of the galaxies, color coded by the timescale $\tau$, from the MCMC run from with the uniform prior for $\tau$. The galaxies with slow rising SFR's are the galaxies above the center of the MS, while galaxies with $\tau$ predicted from @speagleHighlyConsistentFramework2014 are in the lower part of the MS. In the center of the MS are the rest of the galaxies](method_comparison/ms_plots/ms_tau_up.png){#fig-MS-tau}

@@

SFH Reconstructions of the Local Cosmological Volume, based on the three different methods
:::

@@

>>>>

abstract: |
  The Star Formation History (SFH) of a galaxy can offer many insights not only for the evolution and the future of the galaxy, but also for the evolution of the Universe. This is why there are various theoretical models trying to describe the SFH of galaxies. One of those models is the Delayed-Tau model, which approximates the star formation of galaxies as a function with a rising Star Formation Rate (SFR) at the beginning, until it reaches a peak at a time $\tau$, unique for each galaxy, and then it drops at an exponential rate.

  @haslbauerCosmologicalStarFormation2023 argue that the use of Delayed-$\tau$ model for the galaxies of the Local Cosmological Volume (LCV) is opposed to the Lilly-Madau plot ([@madauCosmicStarFormation2014]), which plots the observed SFR's of galaxies with the corresponding redshifts ($z$) and calculates a cosmic SFR peak at $z\approx 2$. The way they calculated this inconsistency is by using observational data for SFR and galaxy masses from the UNGC catalog (@karachentsevUPDATEDNEARBYGALAXY2013, @karachentsevSTARFORMATIONPROPERTIES2013) for calculating the parameters (the timescale $\tau$ and the normalization constant $A_{del}$) of the model. This calculation for the galaxies of the Local Cosmological Volume (LV), allows the investigation of the SFR throughout the life of each galaxy and so we can find the expected time of peak of the SFR.

@@

# Galaxies and Galaxy Morphology

@@

[^1]: For more information about the Great Debate, visit <https://apod.nasa.gov/debate/1920/cs_real.html>

@@

    -   Assigned positive Numerical Morphological T-Types, extending to +11 for extreme irregulars.

@@

# Star Formation

@@

@@

$$
\int_{t_\text{start}}^{t_\text{final}} \text{SFR}(t)dt = \zeta M_*(t_{sf}),\ t_{sf}=\text{Time of Star Formation,} 
$$ {#eq-sfr_int}

@@

-   **H**\boldmath$\alpha$ **Emission**: Young, hot, massive stars (O-type stars, \~10 Myr, \~20 $M_\odot$) produce a number of ionizing photons, which they ionize the surrounding hydrogen rich gas. The hydrogen undergoes recombination cascades which produce Balmer emission lines of $H\alpha$ (0.6563 $\mu m$) and $H\beta$ (0.4861 $\mu m$). Dust can significantly affect observations. Since this estimator depends on young stars, it gives us information about the present day star formation (\~10 Myr).

-   **Far-Ultraviolet (FUV) Flux**: Mainly emitted by young, hot stars ( B-type stars, \~100 Myr). Dust presence can also significantly affect observations. Since this estimator depends on young stars, it gives us information about the present day star formation (\~100 Myr).

-   **Infrared (IR) Flux**: The stars in a galaxy can heat up the dust in different ways, which then emits radiation in different parts of the IR spectrum. For example, young and massive, short-lived stars, emit UV radiation which then the heated dust emits in a wavelength of $\approx 60 \mu m$, whereas dust heated by UV-faint old or low-mass stars will emit at $\approx 100-150\mu m$. As a result, the total IR emission is age-agnostic and provides a more accurate approximation of the SFR because it accounts for contributions from both young and old stellar populations. Since this estimator is somewhat age agnostic, but is derived from the radiaton of heated dust, it is an estimator for the current star formation rate and not the present SFR (\~ 300 Myr).

@@

where $\mathcal{K}_i$ is a constant specific to each $L_i$ ($i =$ H$\alpha$, IR, radio, FUV, X).

![Plot showing the linear relation $\log\text{SFR}_{FUV} = \log\text{SFR}_{H\alpha}$, as well as their distributions, based on the data from UNGC](figure/log_SFR_FUV_Ha.png){#fig-SFR-Ha-FUV fig-align="center" width="400" fig-pos="t"}

@@

@@

where $\alpha(t)$ and $\beta(t)$ depend on time and redshift $z$ (@speagleHighlyConsistentFramework2014).

![The Main Sequence of the galaxies within the radius of 10 Mpc, with the colorbar representing their morphological type. From the linear fit we get $\alpha(T_0) = 0.86$ and $\beta(T_0)= 9.83$. Early types have lower Stellar Masses and SFR, while spiral galaxies have more intense SF](method_comparison/ms_plots/ms_morph.png){#fig-ms-Type width="60%"}

## Star Formation History Models

@@

![Star Formation Rate over the time of star formation, for different parametric models](images/sfr-models.png){fig-align="center" width="500" fig-pos="ht"}

@@

@@

![The history of cosmic star formation as shown in @madauCosmicStarFormation2014](images/lily-madau.png){fig-align="center" width="450" fig-pos="t"}

@@

![The evolution of the SFR of a galaxy over the time of star formation, according to the delayed-$\tau$. The time need for the galaxy to reach the peak of its star formation is the timescale $\tau$ of the model](images/delayed_tau_sfr.png)

where the current $\text{SFR}_0$ of the galaxies, at the current age of the universe $T_0=13.8$ Gyr, is given in the catalogs. According to these model the timescale, at which the SFR peaks, is around $3.5\lessapprox\tau\lessapprox 4.5$ [@speagleHighlyConsistentFramework2014]. If we intergrate the SFR(t), we can calculate the present day averaged SFR:

@@

$$
\overline{\text{SFR}_\text{del}} = \frac{\zeta M_*}{t_{sf}}
$$ {#eq-averaged-SFR-M}

@@

In our current analysis, we will use two different catalogs of galaxies, containing galaxies within the Local Cosmological Volume (LV), which is defined as a sphere with a radius of 11 Mpc centred around the Milky Way, to remain consistent with with the analysis of @kroupaConstraintsStarFormation2020 and @haslbauerCosmologicalStarFormation2023. The reason we focus on the LV and thus only at small distances is because we want to understand the SFR of the galaxies as they are today $z = 0 \Rightarrow t_\text{emitted}=T_0 = 13.8$ Gyr.

@@

-   **Luminosity and Distance Dependence:** The HECATE catalog is more complete for brighter galaxies (with luminosities above $10^{9.5} L_{B,\odot}$) and distances within 33 Mpc, which includes the LV. For galaxies with luminosities around $10^{10} L_{B,\odot}$, the catalog remains complete up to distances of 100 Mpc.
-   **Star Formation Rate and Stellar Mass:** The percentage of the galaxies with measured SFR is around 50% for distances between 30 and 150 Mpc, primarily due to limitations in the WISE survey coverage. Similarly, the completeness of stellar mass estimates is comparable to that of the B-band luminosity, with overdensities at short distances and a sharp cutoff at larger distances.

The *UNGC* catalog focuses on galaxies within the LV, specifically those with radial velocities less than 600 km/s and distances within 11 Mpc. It is around $40\%-60\%$ complete for galaxies with B-band magnitude $M_B^c < -11^m$. In the Local Group ($D \leq 10 \, \text{Mpc}$) only half the galaxies have $M_B^c > -11^m$ and it is expected that for distances $\sim 9 \, \text{Mpc}$ more than half of the ultra-faint galaxies are missing. Specifically, it is estimated that around $10^3–10^4$ galaxies of the LV are not included in UNG.

@@

Since convergence can not occur for $x\ll 1$, this method should not be used for galaxies in the begining phases of their star formation [^10]. If the assumption, $t_{sf}=13.6$ Gyr for all the galaxies is correct, then an inconstancy in the plot $A_{del}-\text{sSFR}$ should only be observed, for higher values of the sSFR, since, according to the delayed- $\tau$ model, higher sSFR values indicate a rising SFR, while lower SFR values indicate that the bulk of $M_*$ is already produced and the galaxy is in the declining SFR phase. However, as seen from @fig-sSFR-A, this inconsistency occurs in galaxies for both high and low sSFR's, meaning that the the model has convergence issues both for slow rising SFR and quenching galaxies, indicating an intrinsic problem with this approach.

@@

[^11]: For more information about the usage of Stan and its metrics can be found here: <https://mc-stan.org/>

::: {#fig-distr-comparison layout-ncol="2" fig-pos="h"}
![Distribution of $\tau$ values.](method_comparison/tau_hist.png){#fig-tau-hist-both}

@@

Using these priors, the metrics of the Stan software[^11] give good results ($\hat{R}<1.001$ and $\text{ESS} >1000$), while the calculated values correspond to the data of the catalog. Specifically, we use the data to recalculate the SFR from the predicted parameters, using the delayed-$\tau$ model and compare the normalization constant to the stellar masses of the galaxies (@fig-A-M-mcmc).

@@

![Main Sequence of of the galaxies, color coded by the timescale $\tau$, from the MCMC run from with the uniform prior for $\tau$. The galaxies with slow rising SFR's are the galaxies above the center of the MS, while galaxies with $\tau$ predicted from @speagleHighlyConsistentFramework2014 are in the lower part of the MS. In the center of the MS are the rest of the galaxies](method_comparison/ms_plots/ms_tau_up.png){#fig-MS-tau width="60%" fig-pos="h"}

@@

::: {#fig-A-M-mcmc layout="[[50,50],[1]]" fig-pos="h"}
![Uniform $\tau$ prior](method_comparison/A_M_MCMC_uni.png){#fig-A-M-mcmc-uni}

@@

@@

SFH Reconstructions of the Local Cosmological Volume, based on the three different methods. Ignore the Skew prior (Red), is a proof of concept.
:::

@@

<!-- {{<pagebreak>}} -->
